Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-30 writing/paper-rbm/appendices/appendix.Rmd
revisions

# On the one-to-one correspondence between parameters and moments in the RBM distribution {#appendix-rbm}

@@

@@

```{theorem, label="one", echo=TRUE}
Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent

1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta}_1 =\bm{\theta}_2$.
3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\

>>>>

# Some characterizations of RBM distributions on all nodes through parameter and moment conditions {#appendix-rbm}

@@

```{proposition, label="propone", echo=TRUE}
Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{m},H_1,...,H_{n})$ assumes values in $\{-1,1\}^{m+n}$ and follows the RBM$(m, n, \boldsymbol \theta)$ distribution with parameter $\boldsymbol \theta\in \mathbb{R}^{m+n+m*n}$.

Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+m*n}$ and
let $P_i$ and $\E_i$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent

1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta}_1 =\bm{\theta}_2$.
3. $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.

Additionally, the following three statements are equivalent

1. $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+m*n}$.
3. $V_1,\ldots,V_m, H_1,\ldots,H_n$ are independent random variables assuming $-1/1$ with equal probability $1/2$.

@@

```{theorem, label="one", echo=TRUE}
Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
let $P_i$ and $\E_i$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent

1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots,m$, $j=1,\ldots,n$.
2. $\bm{\theta}_1 =\bm{\theta}_2$.
3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.

### 2018-09-30 writing/paper-rbm/paper.Rmd
revisions

\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$ model expectations and E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ expectations given independence for an RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as$\\theta_{11}$, the dependence parameter, deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. Under the data encoding $-1/1$, the parameter space centered at 0 is also helpful for framing parameter configurations that are undesirably large (i.e., these are naturally parameters that have moved too far away from $\bm{0}$ where the RBM model is anchored to be trivially describable and completely problem-free). In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$, and we may express the model probabilities as
$$
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
$$
for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
\begin{equation}
\label{eq:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} 
for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness. In fact, when the cell probabilities \eqref{eq:0} are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters still can be selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further (and somewhat odd) consequence of the RBM parameterization is then that when fitting the RBM model by likelihood-based methods, we already may know the nature of the answer before we begin: namely, such fitting simply will reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model. There then can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.

For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \eqref{eq:0} before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model further will aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

Ultimately, it is not unquestionably clear that RBM models are useful as generative models. As a concern, without appropriate generative behavior in an RBM model, the uncertainty in estimated model parameters becomes impossible to assess realistically and further, the model can lose useful application to prediction problems (e.g., realistic predictive distributions). In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.  For these reasons, we are skeptical about RBMs as probabilistic conceptualizations for data generation.

>>>>

\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$ model expectations and E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ expectations given independence for an RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as $\\theta_{11}$, the dependence parameter, deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further relations between the $-1/1$ data encoding and the $\boldsymbol \theta$ parameterization of the RBM model class are given in the on-line supplementary materials. Based on this material, Proposition \ref{prp:three} also briefly summarizes how the $-1/1$ data encoding induces some attractive theoretical properties in characterizing RBM models.

\begin{proposition}
\label{prp:three}
Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{\nv},H_1,...,H_{\nh})$ assumes values in $\{-1,1\}^{\nv+\nh}$ and follows the RBM$(\nv, \nh, \boldsymbol \theta)$ distribution \eqref{eq:pmf} with parameter $\boldsymbol \theta\in \mathbb{R}^{\nv+\nh+\nv*\nh}$. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{\nv+\nh+\nv*\nh}$ and let $P_i$ and $\E_i$ denote the full joint probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\setlength{\itemsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots, \nv$, $j=1,\ldots,\nh$.
\item $\bm{\theta}_1 =\bm{\theta}_2$.
\item $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.
\end{enumerate}
Additionally, the following three statements are equivalent
\begin{enumerate}
\setlength{\itemsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for all $i=1,\ldots,\nv$, $j=1,\ldots,\nh$.
\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
\item $V_1,\ldots,V_\nv, H_1,\ldots,H_\nh$ are independent random variables assuming $-1/1$ with equal probability $1/2$.
\end{enumerate}
\end{proposition}

Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. Under the data encoding $-1/1$, the parameter space centered at 0 is also helpful for framing parameter configurations that are undesirably large (i.e., these are naturally parameters that have moved too far away from $\bm{0}$ where the RBM model is anchored to be trivially describable and completely problem-free). In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$, and we may express the marginalized model probabilities as
$$
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
$$
for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
\begin{equation}
\label{eq:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} 
for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11}, \theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness. In fact, when the cell probabilities \eqref{eq:0} are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities \eqref{eq:0} are zero, the corresponding marginalized RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters still can be selected to make the appropriate RBM cell probabilities arbitrarily small. Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf. @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization). A further (and somewhat odd) consequence of the RBM parameterization is then that when fitting the RBM model by likelihood-based methods, we already may know the nature of the answer before we begin: namely, such fitting simply will reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model. There then can be no model refinements or smoothing from the RBM. That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.

For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \eqref{eq:0} before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model further will aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding marginalized RBM cell probabilities. In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example). This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

<!-- Ultimately, it is not unquestionably clear that RBM models are useful as generative models. As a concern, without appropriate generative behavior in an RBM model, the uncertainty in estimated model parameters becomes impossible to assess realistically and further, the model can lose useful application to prediction problems (e.g., realistic predictive distributions). In the case of classification, predictive distributions may not be the ultimate goal. While interpretability may not always be of primary importance, we find that the issues of stability and degeneracy go hand in hand with interpretability, and these should not be ignored. In fact, when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications. For these reasons, we are skeptical about RBMs as probabilistic conceptualizations for data generation. -->

While interpretability of parameters may not always be of primary importance in application of an RBM model (e.g. in some applications to prediction/classification), we find that the issues of stability and degeneracy go hand in hand with those of interpretability, and the former should not be ignored. When S-instability is present, small (imperceptible) differences in training data and corresponding fitted models may lead to greatly different probabilities (and thus substantially different classifications). Ultimately, it is not clear that RBM models are typically useful as generative models for observables. Without appropriate generative behavior in an RBM model, uncertainty in estimated model parameters becomes impossible to realistically assess. Further, the model cannot support other inferences important in many prediction problems (e.g., realistic predictive distributions). For these reasons, we are skeptical about RBMs as widely useful probabilistic constructs, at least in their present incarnation.
